Deep mutational scanning of hemagglutinin helps distinguish the evolutionary fate of human H3N2 influenza virus lineages

### 2018-01-05 paper/manuscript.tex
Reorganize trunkvssidebranch figure
Added caption for H3N2 phylogeny figure

\begin{SCfigure*}[\sidecaptionrelwidth][t]
\centering
\includegraphics[width=11.4cm]{figs/H3N2_phylogeny/H3N2_phylogeny.pdf}
\caption{\label{fig:H3N2_phylogeny}
{\bf Human H3N2 HA phylogeny.}
}
\end{SCfigure*}

@@

\begin{figure*}
\centering
\includegraphics[width=13cm]{figs/dms_overview/dms_overview.pdf}
\caption{\label{fig:dms_overview}
{\bf Deep mutational scanning of the Perth/2009 H3 HA.}
(A) We generated mutant virus libraries using a helper-virus approach~\cite{doud2016accurate}, and passaged them at low MOI to establish a genotype-phenotype linkage and to select for functional HA variants. 
Deep sequencing of the variants before and after selection allowed us to estimate each site's amino-acid preferences.
(B) The experiments were performed in full biological triplicate. 
We also passaged and deep sequenced library 3 in duplicate.
(C) Frequencies of nonsynonymous, stop, and synonymous mutations in the mutant plasmid DNA, the passaged mutant viruses, and wildtype DNA and virus controls. 
(D) The Pearson correlations among the amino-acid preferences estimated in each replicate. 
}
\end{figure*}

@@

\begin{figure*}
\centering
\includegraphics[width=16cm]{figs/mut_tolerance/entropy_heatmap.pdf}
\caption{\label{fig:mut_tolerance}
{\bf Mutational tolerance mapped onto each site of HA.}
Mutational tolerance as calculated by the Shannon entropy of a given site's amino-acid preferences are mapped onto the structure of the H3 trimer (PDB 4O5N;~\cite{lee2014receptor}) and the H1 trimer (PDB 1RVX;~\cite{gamblin2004structure}), with both trimers in approximately the same orientation. 
The site entropies were calculated from the preferences measured in the Perth/2009 H3 (left panel) from this study, or the preferences measured in the WSN/1933 H1 (right panel) from~\cite{doud2016accurate}. 
Lighter shades of purple or brown signify low mutational tolerance, while darker shades signify high mutational tolerance. 
For each HA, the structure on the left side colors the full HA trimer, while the structure on the right side colors only one of the monomers.
The sialic acid receptor is shown as black sticks.
The Perth/2009 H3 shows relatively high mutational tolerance in the stalk region, particularly in helix A, compared to the head region. 
High mutational tolerance in H3 was also observed near antigenic site C and near the 190 helix.
The head region of the WSN/1933 H1 is mutationally tolerant compared to the relatively intolerant stalk region. 
}
\end{figure*}

@@

\begin{figure*}
\centering
\includegraphics[width=12cm]{figs/trunkvssidebranch/trunkvssidebranch.pdf}
\caption{\label{fig:trunkvssidebranch}
{\bf Mutations in evolutionary successful strains tend to be more favorable than in strains that die out.}
(A) Phylogenetic tree of human H3N2 influenza virus from 1968-present. 
The trunk is shown in red, and side branches are shown in blue.
The gray branches represent the part of the tree for which we cannot yet distinguish the trunk from side branches.
(B) Using the Perth/2009 H3 preferences, we calculated the log$_{2}$ mutational effect for trunk and side branch mutations in windows of 5 years for every year from 1968-2013. 
The median log$_{2}$ mutational effect in a given window is shown as circles for trunk mutations and triangles for side branch mutations. 
The shaded region demarcates the interquartile range of trunk and side branch mutational effects.
Negative numbers signify mutations towards less preferred amino acids, while positive numbers signify more preferred mutations.
The median trunk mutational effects are consistently higher than the median side branch mutational effects for all windows.
(C) The log$_{2}$ mutational effect for all side branch and all trunk mutations (left panel), in addition to all mutations in internal nodes and terminal nodes on the side branches (right panel) are shown.
The preferences were randomized 10,000 times to estimate significance.
The effects of trunk mutations are higher than side branch internal and terminal node mutations.
}
\end{figure*}

@@

\begin{figure*}
\centering
\includegraphics[width=12cm]{figs/sequence_preference/sequence_preference.pdf}
\caption{\label{fig:sequence_preference}
{\bf The HA sequences of trunk nodes tend to be more preferred than those of side branch nodes.}
We used the preferences to calculate the HA sequence preference normalized by the number of sites of every node in a human H3N2 phylogenetic tree.
To calculate the sequence preference, we used the entire HA sequence (top panel), only epitope sites (center panel) or non-epitope sites (bottom panel) as defined by~\cite{wolf2006long}.
Higher preferences are closer to zero while lower preferences are more negative.
The preferences at all sites and at epitope sites increase as the tree approaches the Perth/2009 strain, indicating epistasis among epitope sites.
The trunk nodes generally exhibit higher sequence preferences than do side branch nodes.
}
\end{figure*}

>>>>

\begin{SCfigure*}[\sidecaptionrelwidth][t]
\centering
\includegraphics[width=11.4cm]{figs/H3N2_phylogeny/H3N2_phylogeny.pdf}
\caption{\label{fig:H3N2_phylogeny}
{\bf Human H3N2 HA phylogeny from 1968-2017.}
The trunk is shown in red, and side branches are shown in blue.
The gray branches represent the part of the tree for which we cannot yet distinguish the trunk from side branches.
The Perth/2009 strain is indicated with a star.
}
\end{SCfigure*}

@@

\begin{SCfigure*}[\sidecaptionrelwidth][t]
\centering
\includegraphics[width=12cm]{figs/dms_overview/dms_overview.pdf}
\caption{\label{fig:dms_overview}
{\bf Deep mutational scanning of the Perth/2009 H3 HA.}
(A) We generated mutant virus libraries using a helper-virus approach~\cite{doud2016accurate}, and passaged them at low MOI to establish a genotype-phenotype linkage and to select for functional HA variants. 
Deep sequencing of the variants before and after selection allowed us to estimate each site's amino-acid preferences.
(B) The experiments were performed in full biological triplicate. 
We also passaged and deep sequenced library 3 in duplicate.
(C) Frequencies of nonsynonymous, stop, and synonymous mutations in the mutant plasmid DNA, the passaged mutant viruses, and wildtype DNA and virus controls. 
(D) The Pearson correlations among the amino-acid preferences estimated in each replicate. 
}
\end{SCfigure*}

@@

\begin{SCfigure*}[\sidecaptionrelwidth][t]
\centering
\includegraphics[width=11.4cm]{figs/mut_tolerance/entropy_heatmap.pdf}
\caption{\label{fig:mut_tolerance}
{\bf Mutational tolerance mapped onto each site of HA.}
Mutational tolerance as calculated by the Shannon entropy of a given site's amino-acid preferences are mapped onto the structure of the H3 trimer (PDB 4O5N;~\cite{lee2014receptor}) and the H1 trimer (PDB 1RVX;~\cite{gamblin2004structure}), with both trimers in approximately the same orientation. 
The site entropies were calculated from the preferences measured in the Perth/2009 H3 (left panel) from this study, or the preferences measured in the WSN/1933 H1 (right panel) from~\cite{doud2016accurate}. 
Lighter shades of purple or brown signify low mutational tolerance, while darker shades signify high mutational tolerance. 
For each HA, the structure on the left side colors the full HA trimer, while the structure on the right side colors only one of the monomers.
The sialic acid receptor is shown as black sticks.
The Perth/2009 H3 shows relatively high mutational tolerance in the stalk region, particularly in helix A, compared to the head region. 
High mutational tolerance in H3 was also observed near antigenic site C and near the 190 helix.
The head region of the WSN/1933 H1 is mutationally tolerant compared to the relatively intolerant stalk region. 
}
\end{SCfigure*}

@@

\begin{SCfigure*}[\sidecaptionrelwidth][t]
\centering
\includegraphics[width=11.4cm]{figs/trunkvssidebranch/trunkvssidebranch.pdf}
\caption{\label{fig:trunkvssidebranch}
{\bf Mutations in evolutionary successful strains tend to be more favorable than in strains that die out.}
(A) Using the Perth/2009 H3 preferences, we calculated the log$_{2}$ mutational effect for trunk and side branch mutations in windows of 5 years for every year from 1968-2013. 
The median log$_{2}$ mutational effect in a given window is shown as circles for trunk mutations and triangles for side branch mutations. 
The shaded region demarcates the interquartile range of trunk and side branch mutational effects.
Negative numbers signify mutations towards less preferred amino acids, while positive numbers signify more preferred mutations.
The median trunk mutational effects are consistently higher than the median side branch mutational effects for all windows.
(B) The sequence preference normalized by the number of sites of every node in the human H3N2 phylogenetic tree shown in Fig.~\ref{fig:H3N2_phylogeny}.
Trunk nodes are in red, side branch nodes in blue, and the Perth/2009 strain is marked with a star.
Higher preferences are closer to zero while lower preferences are more negative.
(C) The log$_{2}$ mutational effect for all side branch and all trunk mutations (left panel), in addition to all mutations in internal nodes and terminal nodes on the side branches (right panel) are shown.
The preferences were randomized 10,000 times to estimate significance.
The effects of trunk mutations are higher than side branch internal and terminal node mutations.
}
\end{SCfigure*}

@@

\begin{figure*}
\centering
\includegraphics[width=12cm]{figs/sequence_preference/sequence_preference.pdf}
\caption{\label{fig:sequence_preference}
{\bf The HA sequences of trunk nodes tend to be more preferred than those of side branch nodes.}
We used the preferences to calculate the HA sequence preference normalized by the number of sites of every node in a human H3N2 phylogenetic tree.
To calculate the sequence preference, we used the entire HA sequence (top panel), only epitope sites (center panel) or non-epitope sites (bottom panel) as defined by~\cite{wolf2006long}.
The preferences at all sites and at epitope sites increase as the tree approaches the Perth/2009 strain, indicating epistasis among epitope sites.
The trunk nodes generally exhibit higher sequence preferences than do side branch nodes.
}
\end{figure*}
